---
title: "The SEIaImIsRD model"
author: "Liangti Dai, David Seiferth and Solveig A. van der Vegt"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{SEIaImIsRD model}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(comomodels)
library(ggplot2)
```

## Introduction
This document introduces the SEIaImIsRD model. This model is an extension of the [SEIRD model](https://github.com/Como-DTC-Collaboration/como-models/blob/main/vignettes/SEIRD.Rmd) where instead of having one population of infected individuals there are three, each with their own level of symptoms. In this model, populations of asymptomatic, mildly symptomatic and severely ill infectious individuals are included. 
The motivation for including these separate populations of infectious individuals is two-fold. First, asymptomatic individuals are much less likely to be detected and recorded as infected than those with symptoms. The count of infected individuals might thus underestimate the actual size of the epidemic, and accounting for this in the model by including a group of asymptomatic individuals could paint a more realistic picture.
Second, individuals with severe symptoms require treatment in hospitals and could die because of the disease, whereas asymptomatic and mildly symptomatic individuals will likely recover without needing clinical care. Splitting the population of infectiousness individuals into these three groups thus allows us to estimate the burden of COVID-19 related clinical care on the healthcare system. Estimating this burden has been a crucial modelling component throughout the COVID-19 pandemic, and estimates from modelling teams have, in part, motivated governments to impose a variety of interventions.

We start by introducing the equations of the model and the computation of the basic reproduction number, $R_0$. We then show how the implementation of this model in the `como-models` package can be used.

## The SEIaImIsRD Model
To build the SEIaImIsRD model with three compartments of infectious individuals, we start with the basic SEIRD model. As a reminder, this model consists of a group of susceptible individuals, $S$, who are not infected but can be if they come in contact with an infectious individual; a group of exposed individuals, $E$, who have been infected but are not infectious yet; a group of infectious individuals, $I$, who can infect susceptibles; a group of recovered individuals, $R$, who are now immune to reinfection; and a group of individuals, $D$, who have died due to COVID-19 infection.

To account for the varying severity of symptoms, we take the group of infectious individuals, $I$, and split them into three groups: asymptomatic individuals, $I_a$; mildly symptomatic individuals, $I_m$; and severely ill individuals, $I_s$. Thus, this model consists of seven ordinary differential equations (ODEs) to track all groups of individuals. The severity of symptoms can affect the probability that an infectious individual dies or infects someone else, for example, an asymptomatic individual will not cough and is thus less likely to transmit their infection.

The SEIaImIsRD model thus consists of the following ODEs:


$$\frac{\text{d}S}{\text{d}t} = -S (\beta_aI_a + \beta_mI_m + \beta_aI_s),$$
$$\frac{\text{d}E}{\text{d}t} = S (\beta_aI_a + \beta_mI_m + \beta_sI_s) - \kappa E,$$
$$\frac{dI_{a}}{dt} = \eta_a \kappa E - (\gamma_a + \mu_a)I_a,$$
$$\frac{dI_{m}}{dt} = \eta_m \kappa E - (\gamma_m + \mu_m)I_m,$$
$$\frac{dI_{s}}{dt} = \eta_s\kappa E - (\gamma_s + \mu_s)I_s,$$
$$\frac{\text{d}R}{\text{d}t} =  \gamma_a  I_a + \gamma_m I_m + \gamma_sI_s,$$
$$\frac{\text{d}D}{\text{d}t} =  \mu_a  I_a + \mu_m I_m + \mu_sI_s,$$
where, for $i\in\{a, m, s\}$, $\beta_i$ are the rates at which an infected individual of symptom group $i$ infects susceptible individuals; $\eta_i$ are the fractions of exposed individuals moving into each of the different infected groups $i$ with $\sum_i\eta_i=1$; $\gamma_i$ are the rates at which individuals of symptom group $i$ recover from the disease; and $\mu_i$, the rates at which individuals of symptom group $i$ die of the disease. The rate of becoming infectious when infected, $\kappa$, is the same for all groups of infectious individuals. All parameters are non-negative.

In addition to these seven main ODEs above, the model also keeps track of the total number of cases ($C$) over time. Here, we define a case to be anyone infected with COVID-19, and thus the rate of appearance of new cases is given by:

$$\frac{\text{d}C}{\text{d}t} = S (\beta_aI_a + \beta_mI_m + \beta_aI_s).$$
The system is closed by specifying the initial conditions
$S(0) = S_\text{init},\ E(0) = E_\text{init},\ I_a(0) = I_{a,\text{init}},\ I_m(0) = I_{m,\text{init}},\ I_s(0) = I_{s,\text{init}},\ R(0) = R_\text{init},\ D(0) = D_\text{init}$. We assume $C_\text{init} = 0$ in the `comomodels` package to count only infections arising throughout a given epidemic.

With the ODEs and initial conditions specified, the system can be solved numerically to obtain the time traces $S(t)$, $E(t)$,$I_a(t)$, $I_m(t)$,$I_s(t)$, $R(t)$, $D(t)$, and $C(t)$. In the `comomodels` package, the populations are normalized so the total population is equal to 1, requiring that $S(t)+E(t)+I_a(t)+I_m(t)+I_s(t)+R(t)+D(t) = 1$ for all $t$.

## Basic reproduction number
Before we numerically solve the SEIaImIsRD model, we derive an expression for the basic reproduction number, $R_0$. The basic reproductive number is a widely used metric in epidemiological modelling and indicates whether a disease outbreak will occur. $R_0$ represents the average number of individuals that an infectious person will infect in a population where everyone is susceptible. If $R_0>1$, each individual typically infects more than one person, and the epidemic will likely grow. If $R_0<1$, the epidemic will likely subside. In this section, we compute the value of $R_0$ for the SEIaImIsRD model. For more mathematical detail on the method used, we refer the interested reader to [@vandenDriessche17].

We calculate the basic reproduction number ($R_0$) using the next generation matrix approach. First, we select those populations in the model that contain infected individuals, i.e. those that are exposed, $E$, or infectious, $I_a$, $I_m$ and $I_s$. We define $x = [E, I_a, I_m, I_s]$ and rewrite the ODEs of the variables in $x$ to be of the form: 
$$\frac{\text{d}x_i}{\text{d}t} = \mathscr{F_i}(x) - \mathscr{V_i}(x),$$ 
where $\mathscr{F_i}(x)$ is the rate of appearance of new infections in compartment $i$, and $\mathscr{V_i}(x)$ represents the rate of all other transitions out of that compartment. For example, if we take $E$, then $\mathscr{F}_E(x) = S (\beta_aI_a + \beta_mI_m + \beta_sI_s)$ and $\mathscr{V}_E(x) = \kappa E$. Considering $I_i$, $\mathscr{F}_i(x) = 0$ because no new infections are epidemiologically ``born'' directly into the infectious compartments, and $\mathscr{V}_i(x) = (\gamma_i + \mu_i)I_i - \eta_i \kappa E$.

This then allows us to compute the matrices $F$ and $V$ for all variables in $x$:
$$F=\frac{\partial \mathscr{F_i}}{\partial x_j}\Bigr|_{x^*}\hspace{5mm}\text{and }\hspace{5mm}V=\frac{\partial \mathscr{V_i}}{\partial x_j}\Bigr|_{x^*},$$
where $x^*$ represents the disease-free equilibrium for this model, where $S=1$. For the SEIaImIsRD model, matrices $F$ and $V$ are:
$$
F = \left(\begin{array}{cc} 
0 & \beta_{a} & \beta_{m} & \beta_{s}\\
0 & 0 & 0 & 0\\
0 & 0 & 0 & 0\\
0 & 0 & 0 & 0
\end{array}\right), 
\hspace{5mm}V = \left(\begin{array}{cc} 
\kappa & 0 & 0 & 0\\
-\kappa \eta_a & \gamma_a + \mu_a & 0 & 0\\
-\kappa \eta_m & 0 & \gamma_m + \mu_m & 0\\
-\kappa \eta_s & 0 & 0 & \gamma_s + \mu_s
\end{array}\right)
$$
where $S_0$ is the initial number of susceptibles, which is equal to one as the whole population is assumed to be susceptible.

The matrix $K=FV^{-1}$ is called the next generation matrix and its $(i,j)$th entry is equal to the expected number of secondary infections in compartment $i$ produced by an infected individual introduced in compartment $j$. For the SEIaImIsRD model, it is given by:

$$
K = \left(\begin{array}{cc} 
\sum_i\eta_i\phi_i & \phi_a & \phi_m & \phi_s\\
0 & 0 & 0 & 0\\
0 & 0 & 0 & 0\\
0 & 0 & 0 & 0
\end{array}\right),\hspace{5mm} \text{where: }
\phi_i = \beta_{i} / (\gamma_i+\mu_i).
$$
The next generation matrix has an intuitive interpretation. Considering element $K_{1,2}=\beta_{a} / (\gamma_a+\mu_a)$, this is the number of new infections that a single asymptomatic individual on average produces. This quantity is equal to the number of contacts per unit time, $\beta$, times the average duration of infectiousness, $1/ (\gamma_a+\mu_a)$. The same holds true for the other infectious compartments. The number of new infected individuals produced on average by a single infected individual is given by the $K_{1,1}$ element: this is given by a weighted average of the numbers produced by each infectious compartment where the weights are the proportions of infected individuals flowing into the various infectious compartments. All rows apart from the first are zero because no new infections are produced that fall directly into the infectious compartments.

The basic reproduction number for the SEIaImIsRD model is then defined as
$$R_0 = \rho(K^{-1}),$$
where $\rho$ denotes the spectral radius of a matrix (i.e. the largest absolute value of its eigenvalues). Due to the sparsity of this matrix, this is given by the $K_{11}$ element:

$$
R_0 = \sum_i\eta_i \phi_i.
$$

## The `SEIaImIsRD` class
In this section we walk through an example of how to use the `SEIaImIsRD` class, the implementation of the SEIaImIsRD model in the `como-models` package.

### Initialization of the model
To create a new `SEIaImIsRD` object, run:
```{r}
model <- SEIaImIsRD()
```

Next, we set the initial conditions. Because the populations are normalized, the populations must sum to one, or an error will be thrown.
```{r}
S <- 0.99
E <- 0.01
asymptomatic <- 0
mild <- 0
severe <- 0
R <- 0
D <- 0

initial_conditions(model) <- list(S = S, E = E, asymptomatic = asymptomatic,
                                  mild = mild, severe = severe, R = R, D = D)
```

We then set the parameter values. As with the other models, we set these parameters based on the COVID-19 pandemic. For a detailed derivation of the parameter values, see the `SEIRD_age` model vignette. We assume here that only individuals with severe symptoms die of the disease, i.e. $\mu_a = \mu_m = 0$. We also assume that asymptomatic and mildly symptomatic individuals recover at approximately the same rate, but severely symptomatic individuals recover slower. Approximately 35% of infected individuals will never develop symptoms, and thus belong to the asymptomatic group [@Sahe21]. The hospitalisation rate has varied significantly over the course of the COVID-19 pandemic but remains on the order of 10 per 100,000 confirmed infection which gives an indication of the fraction of infections that result in severe symptoms [@HospRates]. The fraction of infections with mild infections is then the vast majority: $1 - 0.35 - 0.0001 = 0.6499$. We assume that all infectious individuals are equally infectious. Below, we investigate what happens when we do not make this assumption.
```{r}
beta <- list(asymptomatic = 0.707, mild = 0.707, severe = 0.707)
kappa <- 1/5.5
eta_symptom <- list(mild = 0.6499, severe=0.0001)
gamma <- list(asymptomatic = 0.5, mild = 0.5, severe = 0.29)
mu <- list(asymptomatic = 0, mild = 0, severe = 0.01)

transmission_parameters(model) <- list(beta = beta, kappa = kappa,
                                       p_symptom = eta_symptom, gamma = gamma, mu = mu)
```

To check that the parameters have been set correctly, we call
```{r}
transmission_parameters(model)
```

Lastly, we set time span and time step for the simulation outputs. Here we choose to output results at a daily resolution. Note that the numerical solvers used to integrate the ODEs (by default, the LSODA method) typically compute the variables at many more time points for numerical accuracy. The results are returned only for the specified time points.
```{r}
t <- seq(0, 250, by = 1)
```

We can now simulate the model with the function `run()` and compute $R_0$ for the specified parameter values with the function `R0`. We plot the evolution of $S$, $E$, $I_a$, $I_m$, $I_s$, $R$, and $D$ over time. We also plot the number of cases per day, and the number of deaths per day.
```{r,  fig.align = "center", fig.show="hold",fig.height = 5, fig.width = 7}
par(mfrow=c(1,2))
# run the model
output <- run(model, t)
# compute R0
R0 <- R0(model)
print(paste0("model: R0=", R0))

output_incidences <- output$changes[output$changes$compartment == "Incidence",]
output_deaths <- output$changes[output$changes$compartment == "Deaths",]

# plot groups of individuals over time
ggplot(output$states, aes_string(x = "time", y = "value")) +
  geom_line(aes_string(colour = "compartment")) +
      scale_color_brewer(palette = "Dark2") + 
    theme(legend.position = "bottom", legend.title = element_blank()) +
    labs(x = "time (days)", y = "fraction of the population")

# plot number of cases over time
ggplot(output_incidences, aes_string(x = "time", y = "value")) + 
  geom_bar( stat="identity", position=position_dodge()) +
      scale_fill_brewer(palette = "Dark2") + 
    labs(x = "time (days)", y = "fraction of the population per day") +
  ggtitle('Cases each day')

# plot number of deaths over time
ggplot(output_deaths, aes_string(x = "time", y = "value")) + 
  geom_bar( stat="identity", position=position_dodge()) +
      scale_fill_brewer(palette = "Dark2") + 
    labs(x = "time (days)", y = "fraction of the population per day") +
  ggtitle('Deaths each day')

```

#### Varying the value of $\beta$ 
We consider what happens to the model dynamics if we change the infection rate. For example, what if asymptomatic individuals are significantly less infectious than mildly or severely symptomatic ones? We investigate this by lowering $\beta_a$ from 0.707 to 0.5 so that it is now half the infection rate of infectious individuals with severe symptoms.
```{r, fig.align = "center", fig.show="hold",fig.height = 5, fig.width = 7}
model2 <- model
transmission_parameters(model2) <- list(beta = list(asymptomatic = 0.50, mild =  0.707, severe = 0.707),
                                        kappa = kappa,
                                        p_symptom = eta_symptom, gamma = gamma,
                                        mu = mu)
output <- run(model2, t)
R0 <- R0(model2)

output_incidences <- output$changes[output$changes$compartment == "Incidence",]
output_deaths <- output$changes[output$changes$compartment == "Deaths",]

# plot groups of individuals over time
ggplot(output$states, aes_string(x = "time", y = "value")) +
  geom_line(aes_string(colour = "compartment")) +
      scale_color_brewer(palette = "Dark2") + 
    theme(legend.position = "bottom", legend.title = element_blank()) +
    labs(x = "time (days)", y = "fraction of the population")

# plot number of cases over time
ggplot(output_incidences, aes_string(x = "time", y = "value")) + 
  geom_bar( stat="identity", position=position_dodge()) +
      scale_fill_brewer(palette = "Dark2") + 
    labs(x = "time (days)", y = "fraction of the population per day") +
  ggtitle('Cases each day')

# plot number of deaths over time
ggplot(output_deaths, aes_string(x = "time", y = "value")) + 
  geom_bar( stat="identity", position=position_dodge()) +
      scale_fill_brewer(palette = "Dark2") + 
    labs(x = "time (days)", y = "fraction of the population per day") +
  ggtitle('Deaths each day')

```
We see that when $\beta_a$ is lowered, a far smaller fraction of the population gets infected. This is what we would expect. When we now compute $R_0$ for the model with this new parameter set, we get
```{r}
print(paste0("model2: R0=", R0))
```
So although $R_0>1$ still, its value is smaller than before, so the epidemic develops less quickly. This is to be expected if part of the infectious population is less infectious.

## References