---
title: "The SEIRD model"
author: "Hui Jia Farm and Solveig A. van der Vegt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The basic SEIR model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(comomodels)
library(ggplot2)
library(glue)
library(dplyr)
```

## Introduction
This document explains the basics of the SEIR model. The model describes
how populations of susceptible, exposed, infectious and recovered
individuals evolve over time. A simple
ordinary differential equation (ODE) is defined for each of the populations. In `comomodels`, we provide users with a premade SEIR model and here use it to illustrate how this type of model works.

## The SEIRD Model
The SEIRD model consists of four ODEs describing four populations of susceptible,
recovered, infectious and recovered individuals. Susceptible
individuals ($S$) have not been exposed to the virus. Exposed individuals ($E$) have been
exposed to the virus and have been infected but are not yet infectious to others; they are 
in the incubation period. Infectious individuals ($I$) can spread the virus to 
others. Recovered individuals ($R$) are no longer
infectious, and, in this model, have acquired immunity as they do not return
to the susceptible population. Births and deaths unrelated to the infection are not considered in this model.
The ODEs describing how the four populations evolve over time are
$$\frac{\text{d}S}{\text{d}t} = -\beta S I,$$
$$\frac{\text{d}E}{\text{d}t} = \beta S I -\kappa E,$$
$$\frac{\text{d}I}{\text{d}t} = \kappa E - (\gamma + \mu) I,$$
$$\frac{\text{d}R}{\text{d}t} = \gamma I,$$
where $\beta$, $\kappa$, $\gamma$ and $\mu$ are positive parameters.
The rate at which susceptible individuals are infected depends on the fraction
of the population that is susceptible, the fraction of the population that is
infectious, and the probability of an infectious individual infecting a 
susceptible individual, $\beta$. This is represented in the
first terms on the right-hand side of $\text{d}S/ \text{d}t$ and $\text{d}E/ \text{d}t$.
Exposed individuals become infectious at rate $\kappa$. $1/ \kappa$ is
the incubation time of the virus, i.e. how many days after exposure the individual
becomes infectious to others. Infectious individuals can either recover or die
because of the disease. Infectious individuals recover in $1/ \gamma$
days, thus individuals recover at rate $\gamma$. Similarly, infectious
individuals die due to the infection after $1/ \mu$ days, making the death rate $\mu$.

In addition to these four main ODEs, the model also keeps track of the cumulative number 
of cases ($C$) and the cumulative number of disease-related deaths ($D$) over time
$$\frac{\text{d}C}{\text{d}t} = \beta S I,$$
$$\frac{\text{d}D}{\text{d}t} = \mu I.$$

The system is closed by specifying the initial conditions
$$S(0) = S_0,\ E(0) = E_0,\ I(0) = I_0,\ R(0) = R_0,\ C(0) = C_0,\ D(0) = D_0,$$
where $C_0 = E_0 + I_0 + R_0$ to ensure the cases that occur before $t=0$ are
also counted.
In the implementation of the model in this package, the variables are normalized so 
$S(t) + E(t) + I(t) + R(t) + D(t) \equiv 1$ for any given $t$. This means that each of these state variables represents the fraction of the population in each of these states at a given point in time.

Steady states of the system described above are computed by setting the right-hand sides of the ODEs equal to zero. Taking the equation for the susceptibles:

$$\frac{\text{d}S}{\text{d}t} = -\beta S I = 0.$$

This only has solutions if either $S=0$, in which case there are no individuals left to infect; or $I=0$, in which case there are no infectious individuals, and the disease does not invade the population. Considering all four ODEs, we see that $I=0$ is required to ensure that $\frac{\text{d}R}{\text{d}t} = 0$ and that $E$ must be zero to ensure that $\frac{\text{d}E}{\text{d}t}= 0$. $S$ and $R$ are then free to take any non-negative value
as long as $S + R + D = 1$.

We next illustrate how this model works using functionality available in `comomodels`.

## The `SEIRD` class

We first create an `SEIRD` object that we will use to run simulations.
```{r}
my_model <- SEIRD()
```

Next, we set the parameter values for $\beta$, $\kappa$, $\gamma$ and $\mu$:
```{r}
params <- list(beta=1, kappa=0.9, gamma=0.5, mu=0.1)
transmission_parameters(my_model) <- params
```

To check that the parameters have been set correctly, we call
```{r}
transmission_parameters(my_model)
```

We now need to set the initial conditions for the model. We only set the initial conditions $S(0)$, $E(0)$, $I(0)$ and $R(0)$.
```{r}
initial_conditions(my_model) <- list(S0=0.99, E0=0, I0=0.01, R0=0)
```
The initial conditions must sum to one. If they do not, an error is thrown.

Now we simulate the system from $t=0$ days to $t=50$ days.
```{r}
times <- seq(0, 50, by = 1)
out_df <- run(my_model, times)
```

The simulation returns two objects: one is a data frame comprising the states over time. Here, there's a peak in the infectious proportion between $t=15$ and $t=20$.
```{r, fig.height = 5, fig.width = 7}
states <- out_df$states

ggplot(states, aes(x = time, y = value)) +
  geom_line(aes(color = compartment)) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "time (days)", y = "fraction of the population") +
  theme(text = element_text(size = 20))    
```

The simulation also outputs incidence and deaths. These are derived from the state variables. Incidence is just the difference between the cumulative number of cases between two time points,

$$ \text{incidence(t)} = C(t) - C(t - 1),$$

since this shows the number of cases which have arisen between these time  points. Deaths are also given by a difference, but, in this instance, between the cumulative number of deaths at two consecutive time points:

$$ \text{deaths(t)} = D(t) - D(t - 1).$$

```{r fig2, fig.height = 5, fig.width = 7}
changes <- out_df$changes

ggplot(changes, aes(x = time, y = value, fill = compartment)) +
  geom_bar(stat="identity", position = position_dodge()) +
  labs(x = "time (days)", y = "fraction of the population per day") +
  theme(legend.position = "bottom", legend.title = element_blank(),
        text = element_text(size = 20))
```

## Sensitivity analysis
In this section, we investigate the sensitivity of the number of deaths to the transmission rate, $\beta$. We run the model for varying values of $\beta$ and plot the
number of deaths per day over time.
```{r, fig.height = 5, fig.width = 7}
# function to setup and run model for different beta values
run_seird <- function(beta_val) {
  inits <- list(S0=0.99, E0=0, I0=0.01, R0=0)
  params <- list(beta=beta_val, kappa=0.9, gamma=0.5, mu=0.1)
  model <- SEIRD(initial_conditions = inits,
                transmission_parameters=params)
  times <- seq(0, 50, by = 1)
  out_df <- run(model, times)
  out_df$changes
}

# run model across different beta values
beta_vals <- c(0.1, 0.5, 0.75, 1, 2)
for(i in seq_along(beta_vals)) {
  beta_temp <- beta_vals[i]
  temp <- run_seird(beta_temp) %>% mutate(beta=beta_temp)
  if(i == 1)
    result <- temp %>% mutate(beta=beta_temp)
  else
    result <- result %>% bind_rows(temp)
}

# plot results
result %>% 
  filter(compartment=="Deaths") %>% 
  ggplot(aes(x=time, y=value, fill = as.factor(beta))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "time (days)", y = "fraction of the population that died per day") +
  theme(legend.position = "bottom", legend.title = element_blank(),
        text=element_text(size=20))

```

Why do some epidemics die out quickly but others don't? This is can be explored through the basic reproductive number, $R_0$, which is the expected number of people infected by a
single infectious individual in a population where everyone is susceptible (i.e $S_0 \approx 1$). Note that this basic reproductive number is unrelated to the initial condition for the recovered population! If $R_0>1$, an infectious individual infects more than one other person and the epidemic initially grows exponentially; if $R_0<1$, an infectious individual infects less than one other person and the epidemic dies out quickly. In an SEIRD model with normalized populations, the reproductive number is given by:

$$R_0 = \beta S_0/(\gamma + \mu).$$
We now calculate $R_0$ for $\beta=0.1$ and $\beta=0.75$.
```{r, fig.height = 5, fig.width = 7}
inits <- list(S0=0.99, E0=0, I0=0.01, R0=0)

beta_val <- 0.1
params <- list(beta=beta_val, kappa=0.9, gamma=0.5, mu=0.1)
model1 <- SEIRD(initial_conditions = inits,
              transmission_parameters=params)

beta_val <- 0.75
params <- list(beta=beta_val, kappa=0.9, gamma=0.5, mu=0.1)
model2 <- SEIRD(initial_conditions = inits,
              transmission_parameters=params)
paste0("R0(beta=0.1) = ", R0(model1))
paste0("R0(beta=0.75) = ", R0(model2))
```

We see that for $\beta = 0.1$, $R_0 < 1$ and for $\beta = 0.75$, $R_0 > 1$. This explains why in the first case the epidemic died out quickly, and in the second case the
infection initially spreads fast!