---
title: "SEIRD_contact_uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEIRD_contact_uncertainty}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(comomodels)
library(tidyverse)
library(ggplot2)
library(socialmixr)
load(file = "../data/population.rda")
```

## Introduction

Age-structured compartmental models such as the SEIRD implemented in this
repository use contact matrices to specify the spread of a disease within and
between age groups. Given a contact matrix $C$, each element $C_{i,j}$
indicates the expected number of contacts someone from age group $i$ has per
day with people from age group $j$.

Typically, when performing simulations of the age-structured SEIR model, or
inference for the parameters of the model, the contact matrix is provided as a
fixed input. However, using fixed values for the contact matrix neglects the
uncertainty which may be present in the contact data.

The purpose of this document is to investigate the sensitivity of the outputs
of the SEIR model to the values of the contact matrix. Using bootstrap samples
which represent the uncertainty in the contact matrix, we show significant
uncertainty in the numbers of infected individuals.

For details and explanation of the SEIR model, see the following notebook:
https://github.com/Como-DTC-Collaboration/como-models/blob/main/vignettes/SEIRD_age_structured.Rmd

## Bootstrap samples of contact matrices

We obtain samples of the contact matrix using the socialmixr library which
accesses the POLYMOD data. This library allows us to generate bootstrap samples
of the contact matrix for countries covered by the study. The first step is to
generate 200 of these samples for the contact matrix in the United Kingdom.

```{r}

# Define age groups and names
ages <- seq(0, 80, 5)
age_names <- vector(length = 16)
for(i in seq_along(age_names)) {
  age_names[i] <- paste0(ages[i], "-", ages[i + 1])
}

# Get population data
pops <- population[population$country == "United Kingdom", ]$pop
pop_fraction <- pops/sum(pops)
pop_fraction[16] <- sum(pop_fraction[16:21])
pop_fraction <- pop_fraction[1:16]
n_ages <- 16

# Load the contact matrix data from POLYMOD
n_bootstrap <- 200
data(polymod)
polymod_data <- contact_matrix(polymod,
                               n=n_bootstrap,
                               countries="United Kingdom",
                               age.limits=ages)

# Get the first element of the list, which contains the matrices
matrices <- polymod_data["matrices"][[1]]

```

First, we inspect the range of values in the sampled contact matrices. In the
plot below, we look at the distribution of the diagonal elements of the matrix
for each age group.

```{r, fig.width=7, fig.height=5}

contacts_same_age <- c()
ages_list <- c()
for (i in 1:n_bootstrap){
  contacts_same_age <- append(contacts_same_age, diag(matrices[[i]][[1]])[1:16])
  ages_list <- append(ages_list, age_names)
}

data <- data.frame(ages_list, contacts_same_age)
data$ages_list <- factor(data$ages_list, levels=age_names[1:16], ordered=TRUE)

ggplot(data, aes(x=ages_list, y=contacts_same_age), l) + geom_boxplot() +
  xlab("Age group") + ylab("Number of contacts with same age group")
```

The plot shows the most uncertainty for ages 5--20 (schoolchildren). In most
of the other age groups, the uncertainty is small but nowhere does it appear 
negligible. In the next section, the effect of these uncertainties on the
SEIR outputs will be studied.

## SEIRD simulations

In this step, we run the age-structured SEIRD model once for each bootstrap
sample of the contact matrix. We use fixed values for the other parameters of
the model. In particular, we set $\beta=2$, $\kappa=1$, and $\gamma=1$, with
an initial exposed group compartment equal to 0.1% of the population size, and
all other individuals in the susceptible group.

The age-structured model takes another parameter $\mu$ which specifies the rate
at which infecteds enter the deceased compartment. For this parameter, we will
provide a vector giving a separate value for each age group. The values are
selected to resemble the death rates of the COVID-19 outbreak in mainland China
during January and February 2020 (Verity, R, et al. "Estimates of the severity
of coronavirus disease 2019: a model-based analysis." The Lancet Infectious
Diseases 20.6 (2020): 669-677.)

```{r}
mu=c(0.000016, 0.000016, 0.00007, 0.00007, 0.00031, 0.00031, 0.0026, 0.0026,
     0.0048, 0.0048, 0.006, 0.006, 0.019, 0.019, 0.043, 0.043)
```

```{r, include=FALSE}
for (i in 1:n_bootstrap){
  matrix=matrices[[i]][[1]]

  # Remove the column and row names so the model will accept it
  colnames(matrix) <- NULL
  rownames(matrix) <- NULL

  # Keep the data for ages 0-80, in 5 year increments
  matrix <- matrix[1:16,1:16]

  model <- comomodels::SEIRDAge(n_age_categories=n_ages,
                   contact_matrix=matrix,
                   age_ranges=as.list(age_names))

  # Set the other parameters of the model
  transmission_parameters(model) <- list(b=2.0, k=1.0, g=1.0, mu=mu)
  initial_conditions(model) <- list(S0=pop_fraction*0.999,
                                    E0=rep(0, n_ages),
                                    I0=pop_fraction*0.001,
                                    R0=rep(0, n_ages),
                                    D0=rep(0, n_ages))
  res <- run(model, time=seq(0, 100, by=1))

  # Save the data for the I and R compartments
  x = filter(res, compartment %in% c("I", "R", "D"))
  if (i==1)
    all_results <- x
  else
    all_results <- rbind(all_results, x)
}
```

Having obtained the simulation results, we plot the central 90% probability
interval of the number in the infected compartment over time for two selected
age groups (15--20 and 75--80).


```{r, fig.width=7, fig.height=5}

I_data <- filter(all_results, age_range=="15-20", compartment=="I")
data <- I_data$value * sum(pops)
dim(data) <- c(length(I_data$time)/n_bootstrap, n_bootstrap)
quants <- t(apply(data, 1, quantile, probs=c(0.05, 0.5, 0.95), na.rm=TRUE))
quants_df <- data.frame(quants)
quants_df["time"] <- seq(0, 100, by=1)

ggplot(quants_df, aes(x = time)) +
  geom_line(aes(y=X5.)) +
  geom_line(aes(y=X50.)) +
  geom_line(aes(y=X95.)) +
  geom_ribbon(aes(ymin=X5., ymax=X95.), fill="blue", alpha=0.5) +
  ylab("I compartment, age 15-20")

I_data <- filter(all_results, age_range=="75-80", compartment=="I")
data <- I_data$value * sum(pops)
dim(data) <- c(length(I_data$time)/n_bootstrap, n_bootstrap)
quants <- t(apply(data, 1, quantile, probs=c(0.05, 0.5, 0.95), na.rm=TRUE))
quants_df <- data.frame(quants)
quants_df["time"] <- seq(0, 100, by=1)

ggplot(quants_df, aes(x = time)) +
  geom_line(aes(y=X5.)) +
  geom_line(aes(y=X50.)) +
  geom_line(aes(y=X95.)) +
  geom_ribbon(aes(ymin=X5., ymax=X95.), fill="blue", alpha=0.5) +
  ylab("I compartment, age 75-80")

```

The results show that the while the shape of the epidemic trajectory remains
similar for all contact matrices in the bootstrap sample, the number of people
in the I compartment exhibits significant uncertainty, particularly near the
peak of the epidemic.

Next, we perform the uncertainty analysis for the number in the recovered
compartment at the final time point, for all age groups.


```{r, fig.width=7, fig.height=5}

data <- filter(all_results, compartment=="R", time==100.0)

ggplot(data, aes(x=age_range, y=value/pop_fraction), l) +
  geom_boxplot() +
  xlab("Age group") +
  ylab("Proportion of age group ever infected")

```
Finally, we study the effects on death.

```{r, fig.width=7, fig.height=5}

data <- filter(all_results, compartment=="D", time==100.0)

ggplot(data, aes(x=age_range, y=value/pop_fraction), l) +
  geom_boxplot() +
  xlab("Age group") +
  ylab("Proportion of age group died")

```
