---
title: "The SEIRD model"
author: "Hui Jia Farm, Solveig A. van der Vegt and Ben Lambert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The basic SEIR model}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, include=FALSE}
library(comomodels)
library(ggplot2)
library(glue)
library(dplyr)
```

## Introduction
This document introduces the deterministic SEIRD model, which is the foundation around which many, more advanced models of COVID-19 transmission dynamics are based (for example, all the other models within this package). In this vignette, we explain what this model represents, and we show how it can be used to understand epidemic dynamics typical of the "waves" of infection seen during the COVID-19 pandemic.

## The SEIRD Model
The SEIRD model is an example of a compartmental model in mathematical epidemiology (see, for example, Anderson and May, 1992). In these types of models, individuals are assumed at a particular point in time to belong to only one of a number of "compartments": discrete containers where all individuals within them are assumed to be in the same state. In an SEIRD model, there are five such compartments / states: Susceptible, Exposed, Infectious, Recovered and Dead. Susceptible individuals ($S$) have not been exposed to the virus. Exposed individuals ($E$) have been exposed to the virus and have been infected but are not yet infectious to others -- they are in the incubation period. Infectious individuals ($I$) can spread the virus to others. Recovered individuals ($R$) are no longer infectious, and, in this model, have acquired immunity as they do not return to the susceptible population. Some individuals, however, unfortunately do not recover from infection and die as a result, falling into the dead state ($D$).

By assuming that individuals belong only to one of these states, the model effectively assumes that all other information outside of these states is unimportant for transmission. So, an individual's age, their location and any pre-existing medical conditions they may have are, for example, unnaccounted for in this model. In COVID-19 transmission, this assumption is unlikely to hold, since we know that individuals' age does, for instance, strongly influence their susceptibility to death (Verity et al., 2020). The SEIRD model can, however, still be used to approximate the aggregate dynamics of the population since it captures the dominant features of human-to-human epidemic dynamics.

In compartmental models, individuals can move between these states following one of a predefined set of paths. In the SEIRD model, susceptible individuals can become infected with the virus after exposure to it from an infectious individual. In doing so, they can move from the susceptible compartment to the exposed, as is illustrated in the left-hand nodes of the below graph as indicated by the $S\rightarrow E$ transition.
```{r, echo=FALSE, warning=FALSE, out.width = "400px"}
library(DiagrammeR)
grViz("
digraph PrimC{
  graph [rankdir = 'LR']
  node [shape = circle]
  S E I R D
  S -> E [label = '&#946; * S * I']
  E -> I [label = 'kappa * E']
  I -> R [label = 'gamma * I']
  I -> D [label = 'mu * I']
  }")
```

Exposed ($E$) individuals then become infectious ($I$) over time as they transition out of the latency period. For infectious individuals, there are two potential transitions: either an individual recovers and they enter the recovered state ($R$) where they remain forever; or they die, entering the dead state ($D$).

In the model that we discuss here, it is assumed that the population size is sufficiently large that we can view the number of individuals in a given compartment on a continuous scale. We also assume that the process governing the dynamics is deterministic: that is, given the current state of the system (i.e. the numbers of individuals in the $S$, $E$, $I$, $R$ and $D$ compartments), we can perfectly forecast its future behaviour. In reality, this is unlikely to hold since our model neglects to consider a myriad of factors that can influence the resultant epidemic. Indeed, stochastic models try to account for these unobserved elements by assuming randomness plays a role in the underlying dynamics. Nevertheless, in many instances, particularly when a population size is large, assuming deterministic system dynamics may give a good approximation of the mean of equivalent stochastic systems (Erban et al.).

Additionally, the modelling framework that we use assumes that the system changes continuously over time. Specifically, the system assumes that changes in the compartments are governed by a system of coupled ordinary differential equations (ODEs). An example of such an equation is:

$$\frac{\text{d}S}{\text{d}t} = -\text{newly exposed} = -\beta S I,$$

where the left-hand side gives the rate of growth of the susceptible compartment. Since, here, we have not modelled births nor natural (i.e. non-COVID-19 deaths), the only way in which the number of susceptible individuals changes is through those individuals becoming exposed to infection. In this model, we assume that individuals are "well-mixed" -- that is, that a susceptible individual can potentially contact with any infected individual in the population: this means that there are $S * I$ contacts between susceptible and infected individuals at a point in time. Note, this is unlikely to be the case in reality, since populations are highly structured, both geographically and societally, meaning that individuals interact mostly with people in their immediate social group. In the model, the rate at which contacts between a susceptible and infected individual results in transmission is given by $\beta>0$, meaning that the overall rate that susceptibles become exposed is given by $\beta * S * I$. Because this leads to a decline in the susceptible group size, we require a minus sign in the above equation for $\frac{\text{d}S}{\text{d}t}$.

Note that $\beta$ is a rate (per unit time) at which a susceptible individual becomes infected on contact with an infected individual. It is not a probability as it is sometimes mistakenly labelled and so $\beta > 1$.

The full SEIRD model consists of four ODEs that a coupled together: the first of these governs the behaviour of susceptible individuals, which has already been introduced. The next equation governs the growth of the exposed compartment:

$$\frac{\text{d}E}{\text{d}t} = \text{newly exposed} - \text{newly infectious} = \beta S I -\kappa E,$$
where the rate of change is given by the difference between the flow into this compartment due to those individuals who have been newly exposed and the flow out of it, through those who have newly become infectious. (These represent the two arrows going into and out of the $E$ compartment in the arrow graph above.)

We can gain insight into how the flow out of the $E$ compartment works by imagining that we run out of infected individuals at a given time: here, we assume $t=0$. The new equation then becomes:

$$\frac{\text{d}E}{\text{d}t} = -\kappa E,$$
Supposing there is a given number of infected individuals at $t=0$ represented by $E=E_0$, we can solve the above equation to yield an expression for the number of exposed over time:

$$E(t) = E_0 \exp (- \kappa t)$$
representing exponential decay, which we can plot assuming $E_0=1$ and $kappa=0.5$:
```{r, fig.height = 5, fig.width = 7}
# function to return exponential decay
E <- function(E_0, kappa, t) {E_0 * exp(-kappa * t)}

# make values to plot
E_0 <- 1
kappa <- 0.5
t <- seq(0, 10, 0.01)
E_vals <- E(E_0, kappa, t)

tibble(t=t, E=E_vals) %>% 
  ggplot(aes(x=t, y=E)) +
  geom_line() +
  xlab("time (days)")
```

We can also show the effect of changing $\kappa$: unsurprisingly, as $\kappa$ increases, the rate at which individuals become infectious increases and there is faster decay.
```{r, fig.height = 5, fig.width = 7}
kappa_1 <- 0.5
kappa_2 <- 1
kappa_3 <- 2
E_1 <- E(E_0, kappa=kappa_1, t=t)
E_2 <- E(E_0, kappa=kappa_2, t=t)
E_3 <- E(E_0, kappa=kappa_3, t=t)
tibble(t, E=E_1, kappa=kappa_1) %>% 
  bind_rows(tibble(t, E=E_2, kappa=kappa_2)) %>% 
  bind_rows(tibble(t, E=E_3, kappa=kappa_3)) %>% 
  ggplot(aes(x=t, y=E, colour=as.factor(kappa))) +
  geom_line() +
  scale_color_brewer("kappa", palette = "Dark2") +
  xlab("time (days)")
```
An alternative way to view these exponential decay curves is as probability distributions governing the time taken for an exposed individual to become infectious. This is referred to as the latency period. Whilst rigorous proof of this requires considering the equivalent stochastic version of this model, the mean from the deterministic model matches the mean of the stochastic equivalent. As such, the waiting time for an individual to transition from exposed to infectious classes is governed by an exponential distribution. Using repeatedly sampled PCR test data from close contacts of confirmed cases in mainland China, a recent study estimated the latency period of COVID-19 as 5.51 days and a log-normal probability distribution was found to represent that waiting period (Xi et al., Estimating the Latent Period of Coronavirus Disease 2019, 2021). Below we compare the waiting period for the log-normal distribution fit in this study with an exponential distribution with the same latency period.

```{r}
# values derived from Xi et al., Table S2 
mu <- 1.61
sigma <- 0.44

# equivalent decay for exponential model
kappa <- 1 / 5.51

# generate densities across rate of time
t <- seq(0, 20, 0.01)
lnorm_vals <- dlnorm(t, mu, sigma)
exp_vals <- dexp(t, kappa)

tibble(t, value=lnorm_vals, model="log-normal") %>% 
  bind_rows(tibble(t, value=exp_vals, model="exponential")) %>% 
  ggplot(aes(x=t, y=value, colour=model)) +
  geom_line() +
  scale_color_brewer("Model", palette = "Dark2") +
  ylab("density") +
  xlab("time (days)")
```

Although these share a common mean waiting period, there are considerable differences between the curves that characterise the waiting period. Show the CDFs

In particular, the exponential curve predicts that $X$ individuals will have become 


Natural births and deaths -- in other words, those unrelated to infection -- are 
not considered in this model. Therefore, the SEIRD model is suitable only to  diseases that span a short period of time, when natural births and deaths  do not significantly affect dynamics. The ODEs describing how the four populations  evolve over time are

$$\frac{\text{d}S}{\text{d}t} = -\beta S I,$$
$$\frac{\text{d}E}{\text{d}t} = \beta S I -\kappa E,$$
$$\frac{\text{d}I}{\text{d}t} = \kappa E - (\gamma + \mu) I,$$
$$\frac{\text{d}R}{\text{d}t} = \gamma I,$$
where $\beta>0$, $\kappa>0$, $\gamma>0$ and $\mu>0$.
The rate at which susceptible individuals are infected depends on the fraction
of the population that are susceptible and infectious and the rate at which interactions between infectious and susceptible individuals lead to infections, $\beta$.
Exposed individuals become infectious at rate $\kappa$. $1/ \kappa$ is
the mean incubation time of the virus, i.e. how many days after exposure the individual
becomes infectious to others. Infectious individuals can either recover or die
because of the disease. Infectious individuals recover at a mean of $1/ \gamma$
days. Similarly, infectious
individuals die due to the infection after a mean of $1/ \mu$ days.

In addition to these four main ODEs, the model also keeps track of the 
cumulative number of cases ($C$) and the cumulative number of disease-related 
deaths ($D$) over time
$$\frac{\text{d}C}{\text{d}t} = \beta S I,$$
$$\frac{\text{d}D}{\text{d}t} = \mu I.$$

The system is closed by specifying the initial conditions
$$S(0) = S_\text{init},\ E(0) = E_\text{init},\ I(0) = I_\text{init},\ R(0) = R_\text{init},\ C(0) = C_\text{init},\ D(0) = D_\text{init},$$
where $C_\text{init} = E_\text{init} + I_\text{init} + R_\text{init}$ to ensure 
the cases that occur before $t=0$ are also counted.
In the implementation of the model in this package, the variables are normalized 
so$S(t) + E(t) + I(t) + R(t) + D(t) \equiv 1$ for any given $t$. This means that 
each of these state variables represents the fraction of the population in each 
of these states at a given point in time.

Steady states of the system described above are computed by setting the 
right-hand sides of the ODEs equal to zero. Taking the equation for the 
susceptibles:

$$\frac{\text{d}S}{\text{d}t} = -\beta S I = 0.$$

This only has solutions if either $S=0$, in which case there are no individuals 
left to infect; or $I=0$, in which case there are no infectious individuals, and 
the disease does not invade the population. Considering all four ODEs, we see 
that $I=0$ is required to ensure that $\frac{\text{d}R}{\text{d}t} = 0$ and that 
$E$ must be zero to ensure that $\frac{\text{d}E}{\text{d}t}= 0$. $S$ and $R$ 
are then free to take any non-negative value as long as $S + R + D = 1$.

We next illustrate how this model works using functionality available in 
`comomodels`.

## The `SEIRD` class

We first create an `SEIRD` object that we will use to run simulations.
```{r}
my_model <- SEIRD()
```

Next, we set the parameter values for $\beta$, $\kappa$, $\gamma$ and $\mu$:
```{r}
params <- list(beta=1, kappa=0.9, gamma=0.5, mu=0.1)
transmission_parameters(my_model) <- params
```

To check that the parameters have been set correctly, we call
```{r}
transmission_parameters(my_model)
```

We now need to set the initial conditions for the model. We only set the initial 
conditions $S(0)$, $E(0)$, $I(0)$ and $R(0)$.
```{r}
initial_conditions(my_model) <- list(S0=0.99, E0=0, I0=0.01, R0=0)
```
The initial conditions must sum to one. If they do not, an error is thrown.

Now we simulate the system from $t=0$ days to $t=50$ days.
```{r}
times <- seq(0, 50, by = 1)
out_df <- run(my_model, times)
```

The simulation returns two objects: one is a data frame comprising the states 
over time. Here, there's a peak in the infectious proportion between $t=15$ and 
$t=20$.
```{r, fig.height = 5, fig.width = 7}
states <- out_df$states

ggplot(states, aes(x = time, y = value)) +
  geom_line(aes(color = compartment)) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "time (days)", y = "value of each compartment") +
  theme(legend.position = "bottom", legend.title = element_blank(),
        text = element_text(size = 20))
```

The simulation also outputs incidence and deaths. These are derived from the 
state variables. Incidence is just the difference between the cumulative number 
of cases between two time points,

$$ \text{incidence(t)} = C(t) - C(t - 1),$$

since this shows the number of cases which have arisen between these time 
points. Deaths are also given by a difference, but, in this instance, between 
the cumulative number of deaths at two consecutive time points:

$$ \text{deaths(t)} = D(t) - D(t - 1).$$

```{r fig2, fig.height = 5, fig.width = 7}
changes <- out_df$changes

ggplot(changes, aes(x = time, y = value * 1000, fill = compartment)) +
  geom_bar(stat="identity", position = position_dodge()) +
  labs(x = "time (days)", y = "cases per 1000 of the \n population per day") +
  theme(legend.position = "bottom", legend.title = element_blank(),
        text = element_text(size = 20))
```

## Sensitivity analysis
In this section, we investigate the sensitivity of the number of deaths to the 
transmission rate, $\beta$. We run the model for varying values of $\beta$ and 
plot the number of deaths per day over time.
```{r, fig.height = 5, fig.width = 7}
# function to setup and run model for different beta values
run_seird <- function(beta_val) {
  inits <- list(S0=0.99, E0=0, I0=0.01, R0=0)
  params <- list(beta=beta_val, kappa=0.9, gamma=0.5, mu=0.1)
  model <- SEIRD(initial_conditions = inits,
                transmission_parameters=params)
  times <- seq(0, 50, by = 1)
  out_df <- run(model, times)
  out_df$changes
}

# run model across different beta values
beta_vals <- c(0.1, 0.5, 0.75, 1, 2)
for(i in seq_along(beta_vals)) {
  beta_temp <- beta_vals[i]
  temp <- run_seird(beta_temp) %>% mutate(beta=beta_temp)
  if(i == 1)
    result <- temp %>% mutate(beta=beta_temp)
  else
    result <- result %>% bind_rows(temp)
}

# plot results
result %>% 
  filter(compartment=="Deaths") %>% 
  ggplot(aes(x=time, y=value * 1000, fill = as.factor(beta))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "time (days)", y = "deaths per 1000 of the \n population per day",
       fill = "values of beta") +
  theme(legend.position = "bottom", legend.title = element_text(size = 15),
        text = element_text(size = 20))

```

Why do some epidemics die out quickly but others don't? This is can be explored 
through the basic reproductive number, $R_0$, which is the expected number of 
people infected by a single infectious individual in a population where everyone 
is susceptible (i.e $S_0 \approx 1$). Note that this basic reproductive number 
is unrelated to the initial condition for the recovered population! If $R_0>1$, 
an infectious individual infects more than one other person and the epidemic 
initially grows exponentially; if $R_0<1$, an infectious individual infects less 
than one other person and the epidemic dies out quickly. In an SEIRD model with 
normalized populations, the reproductive number is given by:

$$R_0 = \beta /(\gamma + \mu).$$
We now calculate the reproductive number, $R_0$, for $\beta=0.1$ and $\beta=0.75$.
```{r, fig.height = 5, fig.width = 7}
inits <- list(S0=0.99, E0=0, I0=0.01, R0=0)

beta_val <- 0.1
params <- list(beta=beta_val, kappa=0.9, gamma=0.5, mu=0.1)
model1 <- SEIRD(initial_conditions = inits,
              transmission_parameters=params)

beta_val <- 0.75
params <- list(beta=beta_val, kappa=0.9, gamma=0.5, mu=0.1)
model2 <- SEIRD(initial_conditions = inits,
              transmission_parameters=params)
paste0("R0(beta=0.1) = ", R0(model1))
paste0("R0(beta=0.75) = ", R0(model2))
```

We see that for $\beta = 0.1$, $R_0 < 1$ and for $\beta = 0.75$, $R_0 > 1$. This 
explains why in the first case the epidemic died out quickly, and in the second 
case the infection initially spreads fast!