---
title: "SEmIRD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEmIRD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
This vignette introduces the `SEmIRD` model. This model is an extension of the [SEIRD model](https://github.com/Como-DTC-Collaboration/como-models/blob/main/vignettes/SEIRD.Rmd) where there are three exposed compartments opposed to one. The ODEs describing the dynamics are hence:

$$\frac{\text{d}S}{\text{d}t} = -\beta S I,$$
$$\frac{\text{d}E_1}{\text{d}t} = \beta S I -\kappa E_1,$$
$$\frac{\text{d}E_2}{\text{d}t} = \kappa E_1 -\kappa E_2,$$
$$\frac{\text{d}E_3}{\text{d}t} = \kappa E_2 -\kappa E_3,$$

$$\frac{\text{d}I}{\text{d}t} =\kappa E_3 - (\gamma + \mu) I,$$
$$\frac{\text{d}R}{\text{d}t} = \gamma I,$$
$$\frac{\text{d}D}{\text{d}t} = \mu I.$$

## Rationale
The main reason for having multiple exposed compartments is to allow there to be a non-exponential delay period between an individual becoming infected and turning infectious.

```{r setup}
library(comomodels)
```
